---
title: "Practicum 2.1"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
---

# Hypothesis

I believe we are able to distinguish between defensive and offensive players based on their skill attributes.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 11,
  fig.height = 7
)
```

```{r libraries}
library(tidyverse)
```

# First look on availible data

```{r load-data}
players <- readr::read_csv("D:/Škola/Ing/2.semester/OZNAL/Data/players_22.csv")

dim(players)
glimpse(players)
```

# Response Variable Definition

I decided to choose a binary response variable that evaluates whether a player falls into the `DEFENSIVE` or `OFFENSIVE` class.

```{r inspect-positions}
positions <- players %>%
  pull(club_position) %>%
  str_split(",") %>%
  unlist() %>%
  str_trim() %>%
  unique()

positions
```

After listing all positions, I removed all positions that do not clearly express offensive or defensive nature.

```{r classify-players}
offensive_pos <- c("ST", "CF", "LW", "RW", "CAM")
defensive_pos <- c("CB", "LB", "RB", "LWB", "RWB", "CDM")

players_classified <- players %>%
  mutate(
    player_classification = case_when(
      club_position %in% offensive_pos ~ "offensive",
      club_position %in% defensive_pos ~ "defensive",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(player_classification))

players_classified
count(players_classified, player_classification)
```

Here we filter the dataset, so all ambiguous players are removed.

```{r class-balance}
class_table <- players_classified %>%
  count(player_classification)

class_table

ggplot(
  class_table,
  aes(x = player_classification, y = n, fill = player_classification)
) +
  geom_col(width = 0.6) +
  labs(
    title = "Player count: offensive vs defensive",
    x = "Class",
    y = "Player count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

# Attribute Overview

Visualisation of the number of offensive and defensive players in the filtered dataset.

```{r attribute-overview}
cat(names(players_classified), sep = "\n")

attribute_summary_wide <- players_classified %>%
  summarise(
    across(
      everything(),
      list(
        type = ~ class(.x)[1],
        missing = ~ sum(is.na(.x)),
        unique_values = ~ n_distinct(.x, na.rm = TRUE),
        example = ~ .x %>% na.omit() %>% first() %>% as.character()
      ),
      .names = "{.col}__{.fn}"
    )
  )

attribute_overview <- attribute_summary_wide %>%
  pivot_longer(
    cols = everything(),
    names_to = c("attribute", ".value"),
    names_sep = "__"
  ) %>%
  arrange(attribute)

attribute_overview
```

# Initial Feature Selection

For picking the right predictors, it is mandatory to choose attributes that could contribute to determining a class for each player.

```{r selected-features}
selected_features <- c(
  "attacking_crossing",
  "attacking_finishing",
  "attacking_heading_accuracy",
  "attacking_short_passing",
  "attacking_volleys",
  "defending_marking_awareness",
  "defending_standing_tackle",
  "defending_sliding_tackle",
  "mentality_interceptions",
  "mentality_positioning",
  "mentality_vision",
  "power_long_shots",
  "power_shot_power",
  "power_stamina",
  "power_strength",
  "skill_ball_control",
  "skill_dribbling",
  "skill_long_passing",
  "movement_acceleration",
  "movement_agility",
  "movement_sprint_speed"
)

players_skill_predictors <- players_classified %>%
  select(long_name, player_classification, all_of(selected_features))

players_skill_predictors

glimpse(players_skill_predictors)
summary(players_skill_predictors)
```

# Distribution of Selected Features



```{r selected-features-long}
player_skill_per_row <- players_skill_predictors %>%
  pivot_longer(
    cols = all_of(selected_features),
    names_to = "feature",
    values_to = "value"
  )
```

Histograms show how the selected features are distributed by class.

```{r selected-feature-histograms}
ggplot(
  player_skill_per_row,
  aes(x = value, fill = player_classification)
) +
  geom_histogram(
    alpha = 0.5,
    bins = 20,
    position = "identity"
  ) +
  facet_wrap(~ feature, scales = "free") +
  labs(
    title = "Histograms of selected features by player classification",
    x = "Feature value",
    y = "Count"
  ) +
  theme_minimal()
```

Boxplots help visualise the position and overlap of interquartile ranges between offensive and defensive classes.

```{r selected-feature-boxplots}
ggplot(
  player_skill_per_row,
  aes(x = player_classification, y = value, fill = player_classification)
) +
  geom_boxplot(outlier.alpha = 0.1) +
  facet_wrap(~ feature, scales = "free_y") +
  labs(
    title = "Selected skill predictors by classification",
    x = "Player classification",
    y = "Skill predictor value"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8)
  )
```

# Top Feature Candidates

Top feature candidates were chosen based on the visualised distributions. Secondary features were kept as backups in case some top features had high correlation.

```{r top-and-secondary-features}
top_features <- c(
  "attacking_finishing",
  "attacking_volleys",
  "defending_marking_awareness",
  "defending_sliding_tackle",
  "defending_standing_tackle",
  "mentality_interceptions",
  "mentality_positioning",
  "power_long_shots",
  "power_shot_power",
  "skill_ball_control",
  "skill_dribbling"
)

secondary_features <- c(
  "attacking_short_passing",
  "mentality_vision",
  "movement_acceleration",
  "movement_agility",
  "movement_sprint_speed"
)
```

# Pairwise Relationships of Top Features

The `pairs()` plot gives a direct view of pairwise relationships between the strongest candidate predictors.

```{r top-feature-pairs, fig.width=12, fig.height=12}
pairs_data <- players_classified %>%
  select(player_classification, all_of(top_features))

pairs_colors <- ifelse(
  pairs_data$player_classification == "defensive",
  "tomato",
  "steelblue"
)

pairs(
  pairs_data %>% select(-player_classification),
  col = pairs_colors,
  pch = 19,
  cex = 0.45,
  main = "Pairs plot of top candidate features"
)
```

# Correlation Matrix of Top Features

A correlation matrix is great to check and visualize redundancy and multicollinearity.

```{r top-feature-correlations}
top_feature_correlations <- players_skill_predictors %>%
  select(all_of(top_features)) %>%
  cor(use = "pairwise.complete.obs")

top_feature_correlations
```

```{r top-feature-correlation-heatmap}
top_feature_correlations_long <- top_feature_correlations %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature_1") %>%
  pivot_longer(
    cols = -feature_1,
    names_to = "feature_2",
    values_to = "correlation"
  )

ggplot(
  top_feature_correlations_long,
  aes(x = feature_1, y = feature_2, fill = correlation)
) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), size = 3) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation matrix of top features",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

The correlation heatmap showed many strongly correlating predictors in certain skill groups. To reduce redundancy and avoid severe multicollinearity, it is necessary to choose only one representative predictor from each group.

# Final Features

```{r final-features}
final_features <- c(
  "attacking_finishing",
  "defending_standing_tackle",
  "mentality_interceptions",
  "mentality_positioning",
  "skill_dribbling"
)

players_final_features <- players_classified %>%
  select(player_classification, all_of(final_features))

players_final_features_long <- players_final_features %>%
  pivot_longer(
    cols = all_of(final_features),
    names_to = "feature",
    values_to = "value"
  )
```

# Correlation Matrix of Final Features

```{r final-feature-correlations}
final_feature_correlations <- players_final_features %>%
  select(all_of(final_features)) %>%
  cor(use = "pairwise.complete.obs")

final_feature_correlations
```

```{r final-feature-correlation-heatmap}
final_feature_correlations_long <- final_feature_correlations %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature_1") %>%
  pivot_longer(
    cols = -feature_1,
    names_to = "feature_2",
    values_to = "correlation"
  )

ggplot(
  final_feature_correlations_long,
  aes(x = feature_1, y = feature_2, fill = correlation)
) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), size = 3) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation matrix of final features",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

The final predictors now consist of representative features for skill groups and features that are easily interpretable.

# QQ Plots of Final Features

```{r qq-plots-final-features}
ggplot(players_final_features_long, aes(sample = value)) +
  stat_qq(size = 0.5, alpha = 0.5, color = "#1f4e79") +
  stat_qq_line(color = "#b22222", linewidth = 0.4) +
  facet_wrap(~ feature, scales = "free") +
  labs(
    title = "QQ plots of final features",
    x = "Theoretical quantiles",
    y = "Sample quantiles"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 8)
  )
```

Based on a visual QQ plot normality check, I think models other than LDA or QDA can perform better because a few predictors do not show very good normality.

# Pairwise Relationships of Final Features

```{r final-feature-pairs, fig.width=9, fig.height=9}
final_pairs_data <- players_classified %>%
  select(player_classification, all_of(final_features))

final_pairs_colors <- ifelse(
  final_pairs_data$player_classification == "defensive",
  "tomato",
  "steelblue"
)

pairs(
  final_pairs_data %>% select(-player_classification),
  col = final_pairs_colors,
  pch = 19,
  cex = 0.45,
  main = "Pairs plot of final features"
)
```

# Model Choice

After consideration, the main difference I see between Logistic Regression and SVM is the need to handle non-linear relationships. Based on these pair plots, most features seem to be separable in a mostly linear way, so I think Logistic Regression is the ideal first choice. If there are performance issues, SVM can be tested as a follow-up experiment.
