---
title: 'Elements of AI Practicum 1.2'
subtitle: 'Exploratory Data Analysis and Hypothesis Testing'
output:
  html_document:
    toc: yes
    toc_float: yes
    highlight: kate
    toc_depth: 4
  pdf_document:
    toc: no
version: 2025.2.AI
---

### Goal
Test the hypothesis that left-footed players are engaged in offensive positions more often than right-footed players, and compare their market outcomes.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)

data_path <- c("players_22.csv", file.path("Cvicenie 1", "players_22.csv")) %>%
  purrr::keep(file.exists) %>%
  purrr::pluck(1, .default = NULL)

if (is.null(data_path)) {
  stop("Missing file: players_22.csv")
}

fifa22 <- readr::read_csv(data_path, show_col_types = FALSE)

offensive_positions <- c("LF", "CF", "RF", "LW", "RW", "LS", "ST", "RS")
defensive_positions <- c("LWB", "RWB", "LCB", "CB", "RCB", "LB", "RB")
midfield_positions <- c("LM", "LAM", "CM", "RM", "RAM", "CAM", "CDM", "DM", "LDM", "RDM")

data_pos <- fifa22 %>%
  mutate(primary_pos = stringr::str_trim(stringr::str_split_i(player_positions, ",", 1)),
         pos_group = case_when(
           primary_pos %in% offensive_positions ~ "Offense",
           primary_pos %in% defensive_positions ~ "Defense",
           primary_pos %in% midfield_positions ~ "Midfield",
           primary_pos == "GK" ~ "Goalkeeper",
           TRUE ~ "Other"
         )) %>%
  filter(pos_group != "Other")
```

### 1) Position Composition by Preferred Foot

```{r}
position_summary <- data_pos %>%
  count(preferred_foot, pos_group, name = "count") %>%
  group_by(preferred_foot) %>%
  mutate(proportion = count / sum(count)) %>%
  ungroup()

position_summary
```

```{r fig.height=4.2, fig.width=7}
position_summary %>%
  ggplot(aes(x = preferred_foot, y = count, fill = pos_group)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Offense" = "#f28e2b",
                               "Midfield" = "#4e79a7",
                               "Defense" = "#59a14f",
                               "Goalkeeper" = "#9c755f")) +
  labs(title = "Position distribution by preferred foot",
       x = "Preferred foot", y = "Share within foot group", fill = "Position group") +
  theme_bw()
```

### 2) Statistical Test of the Hypothesis

```{r}
pos_table <- xtabs(~ preferred_foot + pos_group, data = data_pos)
pos_table

chisq_result <- chisq.test(pos_table)
chisq_result
```

Interpretation: distribution differs significantly by preferred foot. In this dataset, left-footed players are not over-represented in offense; they are more concentrated in defensive positions.

### 3) Market Value, Wages and Release Clauses

```{r}
market_summary <- data_pos %>%
  group_by(preferred_foot) %>%
  summarise(n = n(),
            mean_value_eur = mean(value_eur, na.rm = TRUE),
            median_value_eur = median(value_eur, na.rm = TRUE),
            mean_wage_eur = mean(wage_eur, na.rm = TRUE),
            mean_release_clause_eur = mean(release_clause_eur, na.rm = TRUE),
            .groups = "drop")

market_summary
```

```{r}
wilcox_value <- wilcox.test(value_eur ~ preferred_foot, data = data_pos)
wilcox_wage <- wilcox.test(wage_eur ~ preferred_foot, data = data_pos)

wilcox_value
wilcox_wage
```

### 4) Young Players and Potential for Improvement

```{r}
young_players <- data_pos %>%
  filter(age <= 23) %>%
  mutate(growth_gap = potential - overall)

young_summary <- young_players %>%
  group_by(preferred_foot) %>%
  summarise(n = n(),
            mean_potential = mean(potential, na.rm = TRUE),
            mean_overall = mean(overall, na.rm = TRUE),
            mean_growth_gap = mean(growth_gap, na.rm = TRUE),
            .groups = "drop")

young_summary
```

```{r}
test_growth <- t.test(growth_gap ~ preferred_foot, data = young_players)
test_growth
```

### 5) Supporting Visualization

```{r fig.height=4, fig.width=7}
market_long <- data_pos %>%
  select(preferred_foot, value_eur, wage_eur, release_clause_eur) %>%
  pivot_longer(cols = c(value_eur, wage_eur, release_clause_eur),
               names_to = "metric",
               values_to = "amount")

market_long %>%
  ggplot(aes(x = preferred_foot, y = amount, fill = preferred_foot)) +
  geom_boxplot(outlier.alpha = 0.1) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_fill_manual(values = c("Left" = "#3153a2", "Right" = "lightgrey")) +
  labs(title = "Market metrics by preferred foot",
       x = "Preferred foot", y = "EUR") +
  theme_bw() +
  theme(legend.position = "none")
```

### Conclusion
- The hypothesis that left-footed players are more often offensive is not supported in this FIFA22 sample.
- Left-footed players have slightly higher average market metrics (value and wage).
- Young left-footed players have high potential too, but right-footed young players show a slightly larger average `potential - overall` gap.
