---
title: 'Elements of AI Practicum 1.3'
subtitle: 'Cleaning Raw Data (Intermediate)'
output:
  html_document:
    toc: yes
    toc_float: yes
    highlight: kate
    toc_depth: 4
  pdf_document:
    toc: no
version: 2025.2.AI
---

### Goal
Create a reusable data-cleaning pipeline for the messy FIFA21 raw dataset.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)

raw_path <- "fifa21_raw_data.csv"
raw_path <- c(raw_path, file.path("Cvicenie 1", raw_path)) %>%
  purrr::keep(file.exists) %>%
  purrr::pluck(1, .default = "fifa21_raw_data.csv")

if (!file.exists(raw_path)) {
  stop("Missing file: fifa21_raw_data.csv. Put the file into the same folder as this Rmd.")
}

raw_fifa21 <- readr::read_csv(raw_path, show_col_types = FALSE)
```

### 1) Initial Inspection

```{r}
dim(raw_fifa21)
head(names(raw_fifa21), 30)
glimpse(raw_fifa21)
```

### 2) Helper Parsers

```{r}
parse_money_eur <- function(x) {
  x %>%
    str_replace_all("\\n", "") %>%
    str_replace_all("€", "") %>%
    str_trim() %>%
    {
      multiplier <- case_when(
        str_detect(., "M$") ~ 1e6,
        str_detect(., "K$") ~ 1e3,
        TRUE ~ 1
      )
      num <- str_remove(., "[MK]$") %>% as.numeric()
      num * multiplier
    }
}

parse_height_cm <- function(x) {
  x_clean <- str_trim(x)
  feet_inches <- str_match(x_clean, "^(\\d+)'(\\d+)\\\"")
  cm_direct <- str_match(x_clean, "^(\\d+)cm$")

  case_when(
    !is.na(feet_inches[, 1]) ~ as.numeric(feet_inches[, 2]) * 30.48 + as.numeric(feet_inches[, 3]) * 2.54,
    !is.na(cm_direct[, 1]) ~ as.numeric(cm_direct[, 2]),
    TRUE ~ suppressWarnings(as.numeric(x_clean))
  )
}

parse_weight_kg <- function(x) {
  x_clean <- str_trim(x)
  lbs_match <- str_match(x_clean, "^(\\d+)lbs$")
  kg_match <- str_match(x_clean, "^(\\d+)kg$")

  case_when(
    !is.na(lbs_match[, 1]) ~ as.numeric(lbs_match[, 2]) * 0.453592,
    !is.na(kg_match[, 1]) ~ as.numeric(kg_match[, 2]),
    TRUE ~ suppressWarnings(as.numeric(x_clean))
  )
}
```

### 3) Standardize Column Names and Remove Newline Characters

```{r}
clean_names <- names(raw_fifa21) %>%
  str_replace_all("\\n", " ") %>%
  str_replace_all("&", " and ") %>%
  str_replace_all("[^A-Za-z0-9]+", "_") %>%
  str_replace_all("_+", "_") %>%
  str_replace_all("^_|_$", "") %>%
  str_to_lower()

fifa21 <- raw_fifa21
names(fifa21) <- clean_names

fifa21 <- fifa21 %>%
  mutate(across(where(is.character), ~str_replace_all(.x, "\\n", " ") %>% str_squish()))
```

### 4) Convert Money, Height and Weight

```{r}
money_cols <- intersect(c("value", "wage", "release_clause"), names(fifa21))

fifa21 <- fifa21 %>%
  mutate(across(any_of(money_cols), parse_money_eur))

if ("height" %in% names(fifa21)) {
  fifa21 <- fifa21 %>% mutate(height_cm = parse_height_cm(height))
}

if ("weight" %in% names(fifa21)) {
  fifa21 <- fifa21 %>% mutate(weight_kg = parse_weight_kg(weight))
}
```

### 5) Tidy `team_and_contract` and `joined`

```{r}
if ("team_and_contract" %in% names(fifa21)) {
  fifa21 <- fifa21 %>%
    tidyr::separate(team_and_contract,
                    into = c("team_name", "contract_info"),
                    sep = "\\s{2,}|,",
                    extra = "merge",
                    fill = "right")

  fifa21 <- fifa21 %>%
    mutate(contract_start = str_extract(contract_info, "\\d{4}(?= ~)"),
           contract_end = str_extract(contract_info, "(?<=~ )\\d{4}"),
           contract_start = as.integer(contract_start),
           contract_end = as.integer(contract_end))
}

if ("joined" %in% names(fifa21)) {
  fifa21 <- fifa21 %>%
    mutate(joined_date = suppressWarnings(parse_date_time(joined,
                                                          orders = c("b d, Y", "B d, Y", "Y-m-d", "d/m/Y"))))
}
```

### 6) Additional Cleanup

```{r}
fifa21 <- fifa21 %>%
  mutate(across(where(is.character), ~na_if(.x, ""))) %>%
  distinct()

summary_missing <- fifa21 %>%
  summarise(across(everything(), ~sum(is.na(.x)))) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "na_count") %>%
  arrange(desc(na_count))

head(summary_missing, 20)
```

### 7) Save Clean Dataset

```{r}
output_path <- file.path(dirname(raw_path), "fifa21_cleaned.csv")
readr::write_csv(fifa21, output_path)

cat("Cleaned file saved to:", output_path, "\n")
```

### Notes
This solution uses tidyverse functional patterns (`across`, vectorized parsers, conditional mutation) instead of imperative loops, matching the recommended style from tutorials.
