---
title: 'Elements of AI Practicum 1.1'
subtitle: 'Welcome to FIFA'
output:
  html_document:
    toc: yes
    toc_float: yes
    highlight: kate
    toc_depth: 4
  pdf_document:
    toc: no
version: 2025.2.AI
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

data_path <- c("players_22.csv", file.path("Cvicenie 1", "players_22.csv")) %>%
  purrr::keep(file.exists) %>%
  purrr::pluck(1, .default = NULL)

if (is.null(data_path)) {
  stop("Missing file: players_22.csv")
}

fifa22 <- readr::read_csv(data_path, show_col_types = FALSE)
```

We believe that the most natural way to introduce you to data science is by exploring data, and by doing so you will learn something about the domain where data come from. For this task, we have chosen football, or as British and Americans call it, soccer.

### Task 1
Load `players_22.csv` into R.

```{r}
glimpse(fifa22)
```

### Task 2
Determine the number of columns and rows, data labeling and heterogeneity.

```{r}
rows <- nrow(fifa22)
columns <- ncol(fifa22)

rows
columns

column_names <- names(fifa22)
head(column_names, 15)

unique(sapply(fifa22, class))
```

The data are labeled (column names are present) and heterogeneous (mixed column types), so it is not suitable for conversion to a single matrix without losing type information.

### Task 3
Count the number of different column types.

```{r}
col_types <- sapply(fifa22, class)
table(col_types)
```

### Task 4
Determine number of players with complete records.

```{r}
complete_players <- sum(complete.cases(fifa22))
complete_players
```

The count is low because some columns are naturally sparse (`player_tags`, `player_traits`, loan and national-team columns, release clauses), so complete records are rare.

### Task 5
Detect duplicate rows and duplicate columns.

```{r}
dup_rows <- sum(duplicated(fifa22))
dup_cols <- sum(duplicated(as.list(fifa22)))

dup_rows
dup_cols
names(fifa22)[duplicated(as.list(fifa22))]
```

### Task 6
Determine numeric ranges and enum choices.

```{r}
numeric_ranges <- fifa22 %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), list(min = ~min(.x, na.rm = TRUE),
                                      max = ~max(.x, na.rm = TRUE))))

numeric_ranges

sample_enums <- fifa22 %>%
  select(where(~is.character(.x) || is.factor(.x))) %>%
  summarise(across(everything(), ~n_distinct(.x, na.rm = TRUE)))

sample_enums
```

### Task 7
Inspect structure and discuss information layout.

```{r}
head(names(fifa22), 25)
tail(names(fifa22), 40)
```

The first columns are mostly identifiers, biography and club/team metadata. The later columns are mostly technical attributes (0-100 ratings). Non-informative fields for player quality modeling can include URLs and purely administrative IDs.

### Task 8
Find the player with the highest `overall` score.

```{r}
top_player <- fifa22 %>%
  slice_max(overall, n = 1, with_ties = FALSE) %>%
  select(short_name, long_name, age, nationality_name, overall,
         player_positions, club_position, wage_eur, club_name)

top_player
```

### Task 9
Among top 100 players by `overall`, count left/right-footed players and proportions.

```{r}
foot_top100 <- fifa22 %>%
  arrange(desc(overall)) %>%
  slice_head(n = 100) %>%
  count(preferred_foot, name = "count") %>%
  mutate(proportion = count / sum(count))

foot_top100
```

### Task 10
Compare average `overall` of left-footed and right-footed players.


overall_by_foot <- fifa22 %>%
  group_by(preferred_foot) %>%
  summarise(avg_overall = mean(overall, na.rm = TRUE),
            n = n(),
            .groups = "drop")

overall_by_foot

t.test(overall ~ preferred_foot, data = fifa22)

Left-footed players have slightly higher average `overall` in this dataset.

### Task 11
Find the youngest left-footed player with `overall > 80`.

young_elite_left <- fifa22 %>%
  filter(preferred_foot == "Left", overall > 80) %>%
  arrange(age, desc(overall)) %>%
  select(short_name, age, overall, club_name, player_positions) %>%
  slice(1)

young_elite_left

### Task 12
Identify five least-valued left-footed players under 35 with highest potential improvement.

left_value_growth <- fifa22 %>%
  filter(preferred_foot == "Left", age < 35) %>%
  mutate(improvement_potential = potential - overall) %>%
  arrange(value_eur, desc(improvement_potential)) %>%
  select(short_name, age, overall, potential, improvement_potential, value_eur, club_name) %>%
  slice_head(n = 5)

left_value_growth

### Task 13
Analyze additional attributes for left-footed players.

best_left_dribbler <- fifa22 %>%
  filter(preferred_foot == "Left") %>%
  arrange(desc(skill_dribbling), desc(overall)) %>%
  select(short_name, skill_dribbling, overall, club_name) %>%
  slice(1)

left_shooting_85_plus <- fifa22 %>%
  filter(preferred_foot == "Left", shooting >= 85) %>%
  summarise(count = n())

best_left_curve <- fifa22 %>%
  filter(preferred_foot == "Left") %>%
  arrange(desc(skill_curve), desc(overall)) %>%
  select(short_name, skill_curve, overall, club_name) %>%
  slice(1)

extra_left_metrics <- fifa22 %>%
  filter(preferred_foot == "Left") %>%
  summarise(mean_long_passing = mean(skill_long_passing, na.rm = TRUE),
            mean_agility = mean(movement_agility, na.rm = TRUE),
            mean_stamina = mean(power_stamina, na.rm = TRUE))

best_left_dribbler
left_shooting_85_plus
best_left_curve
extra_left_metrics

### Closing Remarks
Practicum 1.1 introduces exploratory analysis, basic inference and simple player profiling with tidyverse.
