---
title: "Practicum 2.2"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
---

# Hypothesis

I hypothesize that a striker’s overall rating can be predicted from a set of selected technical, attacking, movement, and mentality-related attributes.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 11,
  fig.height = 7
)
```

```{r libraries}
library(tidyverse)
```

# First look on availible data

```{r load-data}
players <- readr::read_csv("/Users/lukas/Documents/Skola/Ing./2. semester/OZNAL/Data/players_22.csv")

dim(players)
glimpse(players)
cat(names(players), sep = "\n")
```


# Filtering players based on their position

It is essential to only filter out players that play in attacking positions, as well as define a set of feature candidates that we will be filtering to find out which features are the most determining towards predicting the overall of a player.

```{r}
offensive_pos <- c("ST", "CF", "LW", "RW")

strikers <- players %>%
    filter(club_position %in% offensive_pos)

target <- "overall"

candidate_features <- c(
  "attacking_finishing",
  "attacking_heading_accuracy",
  "attacking_short_passing",
  "attacking_volleys",
  "skill_dribbling",
  "skill_ball_control",
  "skill_long_passing",
  "movement_acceleration",
  "movement_sprint_speed",
  "movement_agility",
  "movement_reactions",
  "movement_balance",
  "power_shot_power",
  "power_jumping",
  "power_stamina",
  "power_strength",
  "power_long_shots",
  "mentality_positioning",
  "mentality_vision",
  "mentality_composure",
  "mentality_aggression"
)

strikers_candidate_features <- strikers %>%
  select(club_position, all_of(target), all_of(candidate_features)) %>%
  drop_na()
```

```{r overall-eda}

summary(strikers_candidate_features$overall)

ggplot(strikers_candidate_features, aes(x = overall)) +
  geom_histogram(binwidth = 1, fill = "#4e79a7", color = "white") +
  labs(
    title = "Histogram of feature 'overall'",
    x = "Value",
    y = "Count"
  ) +
  theme_minimal()

ggplot(strikers_candidate_features, aes(x = "", y = overall)) +
  geom_boxplot(fill = "#f28e2b", width = 0.25) +
  coord_flip() +
  labs(
    title = "Boxplot of feature 'overall'",
    x = NULL,
    y = "overall"
  ) +
  theme_minimal()
```

```{r candidate-features-eda, fig.width=14, fig.height=12}
candidate_long <- strikers_candidate_features %>%
  pivot_longer(
    cols = all_of(candidate_features),
    names_to = "feature",
    values_to = "value"
  )

ggplot(candidate_long, aes(x = value)) +
  geom_histogram(bins = 25, fill = "#4e79a7", color = "white") +
  facet_wrap(~ feature, scales = "free") +
  labs(
    title = "Histograms of candidate features",
    x = "Value",
    y = "Count"
  ) +
  theme_minimal()

ggplot(candidate_long, aes(x = feature, y = value)) +
  geom_boxplot(fill = "#f28e2b", outlier.alpha = 0.35) +
  coord_flip() +
  labs(
    title = "Boxplots of candidate features",
    x = "Feature",
    y = "Value"
  ) +
  theme_minimal()
```

```{r predictor-vs-target-scatter, fig.width=14, fig.height=12}
ggplot(candidate_long, aes(x = value, y = overall)) +
  geom_point(alpha = 0.35, size = 0.9, color = "#4e79a7") +
  geom_smooth(method = "lm", se = FALSE, color = "#d62728", linewidth = 0.5) +
  facet_wrap(~ feature, scales = "free_x") +
  labs(
    title = "Scatterplots of predictors vs target feature 'overall'",
    x = "Predictor value",
    y = "overall"
  ) +
  theme_minimal()
```
